---
title: "Problem Set #4"
author: "Nicole Wood (Team NJ with Jade Jevandofsky and Jessika Viveros"
date: "October 21, 2023"
urlcolor: blue
output: 
  pdf_document:
    df_print: tibble
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
```


In last week's problem set (Week 3, Question 5), you used `tidyverse` functions such as `filter()`, `arrange()`, and `select()` to perform data manipulations. In this problem set, you'll be asked to incorporate the use of pipes (`%>%`) into your code.

You'll also be practicing variable creation in this problem set, using both `mutate()` in combination with `if_else()`, `case_when()`, and `recode()`.


# Question 1: Data manipulation using pipes


1. In the code chunk below, complete the following:

    - Load the `tidyverse` library
    - Use the `load()` and `url()` functions to download the `df_school_all` dataframe from the url: `https://github.com/anyone-can-cook/rclass1/raw/master/data/recruiting/recruit_school_allvars.RData`
      - Each row in `df_school_all` represents a high school (includes both public and private)
      - There are columns (e.g., `visit_by_100751`) indicating the number of times a university visited that high school
      - The variable `total_visits` identifies the number of visits the high school received from all (16) public research universities in this data collection sample

```{r}
r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)
rm(list = ls())
install.packages("tidyverse")
library(tidyverse)

load(url("https://github.com/anyone-can-cook/rclass1/raw/master/data/recruiting/recruit_school_allvars.RData"))
```

2. Use the functions `arrange()`, `select()`, and `head()` to do the following:

    - Sort `df_school_all` descending by `total_visits`
    - Select the following variables: `name`, `state_code`, `city`, `school_type`, `total_visits`, `med_inc`, `pct_white`, `pct_black`, `pct_hispanic`, `pct_asian`, `pct_amerindian`
    - Show the first 10 rows of the dataframe, which represents the top 10 most visited schools by the 16 universities
    
    Complete this in 2 ways: (1) without using pipes (the exact same way you did last week) and (2) using pipes (`%>%`) to complete this task using 1 line of code
    
__Without using pipes__:

```{r}
df_school_all_2 <- arrange(df_school_all, desc(total_visits))
df_school_all_2 <- select(df_school_all_2, name, state_code, city, school_type, total_visits, med_inc, pct_white, pct_black, pct_hispanic, pct_asian, pct_amerindian)
head(df_school_all_2, n = 10)
```

__Using pipes ('%>%')__:

```{r}
df_school_all_3 <- arrange(df_school_all, desc(total_visits)) %>% select(name, state_code, city, school_type, total_visits, med_inc, pct_white, pct_black, pct_hispanic, pct_asian, pct_amerindian)
df_school_all_3 %>% head(n=10)
```

3. Building upon the previous question, use the functions `arrange()`, `select()`, `filter()`, and `head()` to do the following (select same variables as above):  

    (A) Top 10 most visited public high schools in California
    (B) Top 10 most visited private high schools in California
    
    Complete this in 2 ways: (1) without using pipes (the exact same way you did last week) and (2) using pipes (`%>%`) to complete the tasks using 1 line of code each

__Without using pipes__:

```{r}
#Top 10 most visited public high schools in California

df_school_all_public <- arrange(df_school_all, desc(total_visits))
df_school_all_public <- select(df_school_all_public, name, state_code, city, school_type, total_visits, med_inc, pct_white, pct_black, pct_hispanic, pct_asian, pct_amerindian)
df_school_all_public <- filter(df_school_all_public, school_type == "public", state_code == "CA")
head(df_school_all_public, n = 10)

#Top 10 most visited private high schools in California

df_school_all_private <- arrange(df_school_all, desc(total_visits))
df_school_all_private <- select(df_school_all_private, name, state_code, city, school_type, total_visits, med_inc, pct_white, pct_black, pct_hispanic, pct_asian, pct_amerindian)
df_school_all_private <- filter(df_school_all_private, school_type == "private", state_code == "CA")
head(df_school_all_private, n = 10)
```

__Using pipes ('%>%')__:

```{r}
#Top 10 most visited public high schools in California

df_school_all_public2 <- arrange(df_school_all, desc(total_visits)) %>% select(name, state_code, city, school_type, total_visits, med_inc, pct_white, pct_black, pct_hispanic, pct_asian, pct_amerindian) %>% filter(school_type == "public", state_code == "CA")
df_school_all_public2 %>% head(n=10)

#Top 10 most visited private high schools in California

df_school_all_private2 <- arrange(df_school_all, desc(total_visits)) %>% select(name, state_code, city, school_type, total_visits, med_inc, pct_white, pct_black, pct_hispanic, pct_asian, pct_amerindian) %>% filter(school_type == "private", state_code == "CA") 
df_school_all_private2 %>% head(n=10)
```


# Question 2: Variable creation using `tidyverse`'s `mutate()`

Often before creating new "analysis" variables, you may want to investigate the values of "input" variables. Here are some examples of checking variable values using `count()`:

```{r eval=FALSE}
# Counts the total number of observations (i.e., rows) in `df_school_all`
df_school_all %>% count()

# Counts the number of observations that have missing values for the variable `med_inc`
df_school_all %>% filter(is.na(med_inc)) %>% count()

# Frequency count of the variable `school_type`
df_school_all %>% count(school_type)
```



1. Use `mutate()` with `if_else()` to create a 0/1 indicator and then use `count()` to generate the following frequency tables:

    - Create 0/1 indicator called `ca_school` for whether the high school is in California and generate the frequency table for the values of `ca_school`
    - Create 0/1 indicator called `ca_pub_school` for whether the high school is a public school in California and generate the frequency table for the values of `ca_pub_school`

    Note: You do not need to assign/retain the indicator variables in the `df_school_all` dataframe.

```{r}
df_school_all %>% 
  mutate(ca_school = if_else(state_code == "CA", 1, 0)) %>% 
    count(ca_school)

df_school_all %>%
  mutate(ca_pub_school = if_else(state_code == "CA" & school_type == "public", 1, 0)) %>%
    count(ca_pub_school)
  
```


2. Complete the following steps to create an analysis variable using `mutate()` and `if_else()`:

    - First, use `select()` to select the variables `name`, `pct_black`, `pct_hispanic`, `pct_amerindian` from `df_school_all`, and assign the resulting dataframe to `df_race`. You'll be using `df_race` for the remaining bullet points below.
    - Use `filter()`, `is.na()`, and `count()` to investigate whether or not the following variables have missing values: `pct_black`, `pct_hispanic`, `pct_amerindian`
    - Use `mutate()` to create a new variable `pct_bl_hisp_nat` in `df_race` that is the sum of `pct_black`, `pct_hispanic`, and `pct_amerindian`. Remember to assign to `df_race`.
    - Create a 0/1 indicator called `gt50pct_bl_hisp_nat` for whether more than 50% of students identify as black, latinx, or native american and generate a frequency table for the values of `gt50pct_bl_hisp_nat`


```{r}
df_race <- select(df_school_all, name, pct_black, pct_hispanic, pct_amerindian) 

#lines 155 - 162 were written to verify that the total na count is actually 0. 

df_race %>% filter(is.na(pct_black)) %>%
  count()

df_race %>% filter(is.na(pct_hispanic)) %>%
  count()

df_race %>% filter(is.na(pct_amerindian)) %>%
  count()

df_race %>% filter(is.na(pct_black) | is.na(pct_hispanic) | is.na(pct_amerindian)) %>%
  count()

df_race <- mutate(df_race, pct_bl_hisp_nat = pct_black + pct_hispanic + pct_amerindian)

df_race <- mutate(df_race, gt50pct_bl_hisp_nat = if_else(pct_bl_hisp_nat > 50, 1, 0)) 

df_race %>% count(gt50pct_bl_hisp_nat)

```


3. Complete the following steps to create an analysis variable using `mutate()` and `case_when()`:

    - First, use `select()` to select the variables `name` and `state_code` from `df_school_all`, and assign the resulting dataframe to `df_schools`
    - Use `case_when()` to create a new variable in `df_schools` called `region` whose values are:
      - `Northeast`, if `state_code` is in: `'CT', 'ME', 'MA', 'NH', 'RI', 'VT', 'NJ', 'NY', 'PA'`
      - `Midwest`, if `state_code` is in: `'IN', 'IL', 'MI', 'OH', 'WI', 'IA', 'KS', 'MN', 'MO', 'NE', 'ND', 'SD'`
      - `West`, if `state_code` is in: `'AZ', 'CO', 'ID', 'NM', 'MT', 'UT', 'NV', 'WY', 'AK', 'CA', 'HI', 'OR', 'WA'`
      - `South`, if `state_code` is not any of the above states (Hint: Use `TRUE` as the condition to specify default value. You can see an example [here](https://dplyr.tidyverse.org/reference/case_when.html).)
    
```{r}
df_schools <- select(df_school_all, name, state_code)

df_schools <- mutate(df_schools, region = case_when(
  state_code %in% c("CT", "ME", "MA", "NH", "RI", "VT", "NJ", "NY", "PA") ~ "Northeast",
  state_code %in% c("IN", "IL", "MI", "OH", "WI", "IA", "KA", "MN", "MO", "NE", "ND", "SD") ~ "Midwest",
  state_code %in% c("AZ", "CO", "ID", "NM", "MT", "UT", "NV", "WY", "AK", "CA", "HI", "OR", "WA") ~ "West",
  .default = "South"))
    
head(df_schools, n = 10) 
tail(df_schools, n = 10)
```

4. Complete the following steps to recode variables using `mutate()` and `recode()`:

    - In the `df_schools` dataframe, replace the values of the `region` variable as follows:
      - Change `Northeast` to `NE`
      - Change `Midwest` to `MW`
      - Change `West` to `W`
      - Change `South` to `S`
    - In the `df_schools` dataframe, create a new variable `state_name` whose value is:
      - `California`, if `state_code` is `CA`
      - `New York`, if `state_code` is `NY`
      - Choose another state of your choice to recode
      - `Other`, if `state_code` is any other state (Hint: Use `.default` to specify the default value)
    
```{r}
names(df_schools)

df_schools <- mutate(df_schools, region = recode(region,
  "Northeast" = "NE",
  "Midwest" = "MW",
  "West" = "W",
  "South" = "S"))

head(df_schools)

df_schools <- mutate(df_schools, state_name = recode(state_code,
  "CA" = "California",
  "NY" = "New York",
  "TN" = "Tennessee",
  .default = "Other"))

head(df_schools)
```

# Create a GitHub issue   


- Go to the [class repository](https://github.com/anyone-can-cook/rclass1_student_issues_f23/issues) and create a new issue.

- Refer to [rclass1 student issues readme](https://github.com/anyone-can-cook/rclass1_student_issues_f23/blob/main/README.md) for instructions on how to post questions or reflections.

- You are also required to respond to at least one issue posted by another student.

- Paste the url to your issue here: https://github.com/anyone-can-cook/rclass1_student_issues_f23/issues/432 
- Paste the url to the issue you responded to here: https://github.com/anyone-can-cook/rclass1_student_issues_f23/issues/522 

# Knit to pdf and submit problem set  

**Knit to pdf** by clicking the "Knit" button near the top of your RStudio window (icon with blue yarn ball) or drop down and select "Knit to PDF"

- Go to the [class website](https://anyone-can-cook.github.io/rclass1/) and under the "Readings & Assignments" >> "Week 4" tab, click on the "Problem set 4 submission link"  
- Submit both .Rmd and pdf files  
- Use this naming convention "lastname_firstname_ps#" for your .Rmd and pdf files (e.g. jaquette_ozan_ps4.Rmd & jaquette_ozan_ps4.pdf)  

