---
title: "Problem Set #6"
author: "Nicole Wood (in group NJ with Jade Levandofsky and Jessika Viveros)"
date: "November 3, 2023"
urlcolor: blue
output: 
  pdf_document: default
---


# Overview:  

In this problem set, you will be using the **stringr** package (part of tidyverse) to work with strings, and the **lubridate** package for working with dates and times. We will ask you to load Twitter data that is saved as an .Rdata file.  


# Question 1: Working with strings


1. Load the following packages in the code chunk below: `tidyverse` and `lubridate`.

```{r, message=F}
library(tidyverse)
library(lubridate)
```


2. Using `str_c()` and the following objects as input, create the string: `"Roses are red, Violets are blue"`  

    - We encourage you to first sketch out what you want to do on some scratch paper.
    - Recall from the lecture example on "Using str_c() on vectors of different lengths", when multiple vectors of different length are provided in the str_c() function, the elements of shorter vectors are recycled. See below.
    
```{r eval=FALSE}
str_c("@", c("ozanj ", "joebruin ", "josiebruin "), sep = "", collapse = ",")

#[1] "@ozanj ,@joebruin ,@josiebruin "
```
    
    - Now try it yourself.
```{r}
vec_1 <- c("Roses", "Violets")
vec_2 <- c("red", "blue")

str_1 <- "are"

# Write your code here
str_c(vec_1, str_1, vec_2, sep = " ", collapse = ", ")
print(str_c)
```


3. [Pig Latin](https://en.wikipedia.org/wiki/Pig_Latin) is a language game in which the first consonant of each word is moved to the end of the word, then `"ay"` is appended to create a suffix. For example, the word `"Wikipedia"` would become `"Ikipediaway"`.

    - Using `str_c()` and `str_sub()`, turn the given `pig_latin` vector into the string: `"igpay atinlay"`
    - We encourage you to first sketch out what you want to do on some scratch paper.
        - First, think about what the final outcome will look like.
        - Then, think about how you can get there. Play around with the `str_sub()` function. What happens when you include different values in the `str_sub()` function?  
    - this is low-key the trickiest question in the problem set. So if you get stuck, ask a question to your group or github and move on. and come back to it later.

```{r}
pig_latin <- c('pig', 'latin')

# Write your code here

pl1 <- str_sub(pig_latin, start = 2, end = -1)
pl2 <- str_sub(pig_latin, start = 1, end = 1)

str_c(pl1, pl2, "ay", sep = "", collapse = " ")
```


4. Using `str_c()` and `str_sub()`, decode the given `secret_message`. Your output should be a string.  

    - Follow the same logic from above.
    - Sketch out what you want to do on some scratch paper. Break it down step by step. Play around with different values for the `str_sub()` function. 
    
```{r}
secret_message <- c('ollowfay', 'ouryay', 'earthay')

# Write your code here

sm1 <- str_sub(secret_message, start = -3, end = -3)
sm2 <- str_sub(secret_message, start = 1, end = -4)

str_c(sm1, sm2, sep = "", collapse = " ")

str_c
```


# Question 2: Working with Twitter data  


1. You will be using Twitter data we fetched from the following Twitter handles: `UniNoticias`, `FoxNews`, and `CNN`.  

    - This data has been saved as an Rdata file.
    - Use the `load()` and `url()` functions to download the `news_df` dataframe from the url: `https://github.com/anyone-can-cook/rclass1/raw/master/data/twitter/twitter_news.RData`
    - Report the dimensions of the `news_df` data frame (rows and columns). Use the `dim()` function.  


```{r}
load(url("https://github.com/anyone-can-cook/rclass1/raw/master/data/twitter/twitter_news.RData"))

dim(news_df)
```


2. Subset your dataframe `news_df` and create a new dataframe called `news_df2` keeping only the following variables: `user_id`, `status_id`, `created_at`, `screen_name`, `text`, `followers_count`, `profile_expanded_url`.  

    - Note in the following questions we will ask you to create a new column and that means you have to assign `<-` the new changes you are making to the existing dataframe `news_df2`. Ex. `news_df2 <- news_df %>% mutate(newvar = mean(oldvar))`

```{r}
news_df2 <- news_df %>%
  select(user_id, status_id, created_at, screen_name, text, followers_count, profile_expanded_url)

names(news_df2)
```


3. Create a new column in `news_df2` called `text_len` that contains the length of the character variable `text`.

    - What is the `class` and `type` of this new column? Make sure to include your code in the code chunk below.
      - __ANSWER__: The class and type of text_len are both integer. 
    
```{r}
news_df2 <- news_df2 %>%
  mutate(text_len = str_length(text))

head(news_df2)
class(news_df2$text_len)
typeof(news_df2$text_len)
```


4. Create an additional column in `news_df2` called `handle_followers` that stores the twitter handle and the number of followers associated with that twitter handle in a string. For example, the entries in the `handle_followers` column should look like this: `@[twitter_handle] has [number] followers.`

    - What is the `class` and `type` of this new column? Make sure to include your code in the code chunk below.
      - __ANSWER__: The class and type of handle_followers in character. 

```{r}
news_df2 <- news_df2 %>%
  mutate(handle_followers = str_c("@", screen_name, " has ", followers_count, " followers."))

head(news_df2)
class(news_df2$handle_followers)
typeof(news_df2$handle_followers)
```


5. Lastly, create a column in `news_df2` called `short_web` that contains a short version of the `profile_expanded_url` without the `http://www.` part of the url. For example, the entries in that column should look something like this: `nytimes.com`.

```{r}
news_df2 <- news_df2 %>%
  mutate(short_web = str_sub(profile_expanded_url, start = 12, end = -1))

head(news_df2)
tail(news_df2)
```

# Question 3: Working with dates/times


1. Using the column `created_at`, create a new column in `news_df2` called `dt_chr` that is a character version of `created_at`.

    - What is the `class` of the `created_at` and `dt_chr` columns? Make sure to include your code in the code chunk below.
      - __ANSWER__: The class of the created_at column is POSIXct and POSIXt. The class of the dt_chr column is character. 

```{r}
news_df2 <- news_df2 %>%
  mutate(dt_chr = as.character(created_at))

head(news_df2)
class(news_df2$created_at)
class(news_df2$dt_chr)
```


2. Create another column in `news_df2` called `dt_len` that stores the length of `dt_chr`.

```{r}
news_df2 <- news_df2 %>%
  mutate(dt_len = str_length(dt_chr))

head(news_df2)
```


3. Next, create additional columns in `news_df2` for each of the following date/time components:

    a. Create a new column `date_chr` for date (e.g. `2020-03-26`) using the column `dt_chr` and the `str_sub()` function.
    a. Do the same for year `yr_chr` (e.g. `2020`).
    a. Do the same for month `mth_chr` (e.g. `03`).
    a. Do the same for day `day_chr` (e.g. `26`).
    a. Do the same for time `time_chr` (e.g. `22:41:09`).  

```{r}
news_df2 <- news_df2 %>%
  mutate(date_chr = str_sub(dt_chr, start = 1, end = 10),
         yr_chr = str_sub(dt_chr, start = 1, end = 4),
         mth_chr = str_sub(dt_chr, start = 6, end = 7),
         day_chr = str_sub(dt_chr, start = 9, end = 10),
         time_chr = str_sub(dt_chr, start = 12, end = -1))

head(news_df2)

```


4. Using the column we created in the previous question `time_chr`, create additional columns in `news_df2` for the following time components: 

    a. Create a new column `hr_chr` for hour (e.g. `22`) using the column `time_chr` and the `str_sub()` function.
    a. Do the same for minutes `min_chr` (e.g. `41`).
    a. Do the same for seconds `sec_chr` (e.g. `09`).

```{r}
news_df2 <- news_df2 %>%
  mutate(hr_chr = str_sub(time_chr, start = 1, end = 2),
         min_chr = str_sub(time_chr, start = 4, end = 5),
         sec_chr = str_sub(time_chr, start = -2, end = -1))

head(news_df2)
```



5. Now let's get some practice with the `lubridate` package.  

    a. Using the `year()` function from the `lubridate` package, create a new column in `news_df2` called `yr_num` that contains the year (e.g. `2020`) extracted from `date_chr`.
    a. Do the same for month `mth_num`.
    a. Do the same for day `day_num`.
    a. Do the same for hour `hr_num`, but extract from `created_at` column instead of `date_chr`.
    a. Do the same for minutes `min_num`. 
    a. Do the same for seconds `sec_num`.

```{r}
news_df2 <- news_df2 %>%
  mutate(yr_num = year(date_chr),
         mth_num = month(date_chr),
         day_num = day(date_chr),
         hr_num = hour(created_at),
         min_num = minute(created_at),
         sec_num = second(created_at))

head(news_df2)
```



6. Using the **new numeric columns** (e.g. day_num, mth_num) you've created in the previous step, reconstruct the date and datetime columns. Namely, add the following columns to `news_df2`:

    a. Use `make_date()` to create new column called `my_date` that contains the date (year, month, day).
    a. Use `make_datetime()` to create new column called `my_datetime` that contains the datetime (year, month, day, hour, minutes, seconds).
    
    - What is the `class` of your `my_date` and `my_datetime` columns? Make sure to include your code in the code chunk below.
      - __ANSWER__: 

```{r}
news_df2 <- news_df2 %>%
  mutate(my_date = make_date(year = yr_chr, month = mth_chr, day = day_chr),
         my_datetime = make_datetime(yr_num, mth_num, day_num, hr_num, min_num, sec_num))

head(news_df2)
```


# Create a GitHub issue   

- Go to the [class repository](https://github.com/anyone-can-cook/rclass1_student_issues_f23/issues) and create a new issue.

- Refer to [rclass1 student issues readme](https://github.com/anyone-can-cook/rclass1_student_issues_f23/blob/main/README.md) for instructions on how to post questions or reflections.

- You are also required to respond to at least one issue posted by another student.


- Paste the url to your issue here: https://github.com/anyone-can-cook/rclass1_student_issues_f23/issues/723 
- Paste the url to the issue you responded to here: https://github.com/anyone-can-cook/rclass1_student_issues_f23/issues/722 


# Knit to pdf and submit problem set  

**Knit to pdf** by clicking the "Knit" button near the top of your RStudio window (icon with blue yarn ball) or drop down and select "Knit to PDF"

- Go to the [class website](https://anyone-can-cook.github.io/rclass1/) and under the "Readings & Assignments" >> "Week 6" tab, click on the "Problem set 6 submission link"  
- Submit both .Rmd and pdf files  
- Use this naming convention "lastname_firstname_ps#" for your .Rmd and pdf files (e.g. jaquette_ozan_ps6.Rmd & jaquette_ozan_ps6.pdf)

