---
title: 'Problem Set #3'
author: "Nicole Wood (in group NJ with Jade Levandofsky and Jessika Viveros)"
date: "October 14, 2023"
output:
  pdf_document: default
  html_document:
    df_print: paged
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```


The aim of this problem set is to give you practice completing data management tasks associated with filtering/isolating observations, sorting observations, and selecting variables. This can be done using the `filter()`, `arrange()`, and `select()` functions from the `tidyverse` package. Filtering the data can also be done using `base R`'s subsetting operators.

For the following questions, you'll be asked to complete the same task multiple ways based on the `tidyverse` and `base R` approaches. We want you to understand that there are several ways to complete the same task and we want you to practice completing the same task in different ways.

**Note:** For those who know how to use the `%>%` (pipe) operator, please **do not** use it for this assignment. We will go over pipe operators and how to use them later on in the quarter. 

# Question 1: Load and inspect `df_event` dataset


1. In the code chunk below, complete the following:

    - Load the `tidyverse` library
    - Use the `load()` and `url()` functions to download the `df_event` dataframe from the url: `https://github.com/anyone-can-cook/rclass1/raw/master/data/recruiting/recruit_event_somevars.RData`
      - Each row in `df_event` represents a recruiting visit

```{r}
r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)
rm(list = ls())
R.Version()
install.packages("tinytex")
install.packages("tidyverse")
install.packages("Rtools")
library(tidyverse)
load(url("https://github.com/anyone-can-cook/rclass1/raw/master/data/recruiting/recruit_event_somevars.RData"))
str(df_event)
```


2. Inspect the `df_event` dataframe:

    - Use `names()` to identify the column names in the dataframe
    - Use `typeof()` to show the data type of the `event_state` column
    - Use `str()` to show the structure of the `med_inc` column
    - Use `table()` to show the categorical values of the `event_type` column
    
```{r}
names(df_event)
typeof(df_event$event_state)
str(df_event$med_inc)
table(df_event$event_type)
```

# Question 2: Filtering/isolating observations

Filtering can be done using multiple approaches: `tidyverse`'s `filter()` function and `base R`'s subsetting operators. Here is an example of using each method to obtain the total number of recruiting visits to California from the `df_event` dataframe:

```{r, eval=FALSE}
# tidyverse using filter()
nrow(filter(df_event, event_state == 'CA'))

# base R using subsetting operators
nrow(df_event[df_event$event_state == 'CA', ])

```


1. Your turn now! Count the number of recruiting events that satisfy all the following criteria:

    - By the University of Massachusetts-Amherst (`univ_id`: `166629`)
    - An out-of-state public high school (use `event_type`, `event_state`, and `instst`, which is the visiting university's home state)
    - Average median household income is greater than or equal to $100,000 (`med_inc`)
    - Make sure to drop any `NA` values
    
    Use `nrow()` to obtain the count. Do the filtering in the 2 ways below. You should get the same answer.

__tidyverse using `filter()`__:
```{r}
nrow(filter(df_event, univ_id == "166629"))
nrow(filter(df_event, event_type == "public hs", event_state != instst))
nrow(filter(df_event, med_inc >= 100000))

```

__base R using subsetting operators__ (hint: use `which()` to drop `NA`s):
```{r}
nrow(df_event[which(df_event$univ_id == "166629"),])
nrow(df_event[which(df_event$event_type == "public hs" & df_event$event_state != df_event$instst),])
nrow(df_event[which(df_event$med_inc >= 100000),])
```

2. Count the number of recruiting events that satisfy all the following criteria:

    - By the University of South Carolina-Columbia (`univ_id`: `218663`) or by the University of Alabama (`univ_id`: `100751`)
    - And either:
      - An in-state 2-year college visit (use `event_type`, `event_state`, and `instst`, which is the visiting university's home state) OR
      - A zip code with population under 10,000 (use `pop_total`)
    - Make sure to drop any `NA` values
    - Note the [order of precedence](https://www.datamentor.io/r-programming/precedence-associativity/): `&` is higher in priority than `|`

__tidyverse using `filter()`__:
```{r}
nrow(filter(df_event, univ_id == "218663" | univ_id == "100751"))
nrow(filter(df_event, univ_id == "218663" | univ_id == "100751", event_type == "2yr college", event_state == instst))
nrow(filter(df_event, univ_id == "218663" | univ_id == "100751", pop_total < 10000))
nrow(filter(df_event, univ_id == "218663" | univ_id == "100751", ((event_type == "2yr college" & event_state == instst) | (pop_total < 10000))))

#9 events are for in-state 2 year college visits AND a zip code with a total population under 10,000 (reason why the total is slightly different from the component parts. 
```

__base R using subsetting operators__ (hint: use `which()` to drop `NA`s):
```{r}
nrow(df_event[which(df_event$univ_id == "218663" | df_event$univ_id == "100751"),])
nrow(df_event[which((df_event$univ_id == "218663" | df_event$univ_id == "100751") & df_event$event_type == "2yr college" & df_event$event_state == df_event$instst),])
nrow(df_event[which((df_event$univ_id == "218663" | df_event$univ_id == "100751") & df_event$pop_total < 10000),])
nrow(df_event[which((df_event$univ_id == "218663" | df_event$univ_id == "100751") & ((df_event$event_type == "2yr college" & df_event$event_state == df_event$instst) | (df_event$pop_total < 10000))),])
```

# Question 3: Sorting observations


1. Create a new dataframe that contains the events in `df_events` sorted by:

    - Ascending `univ_id`
    - Ascending `event_date`
    - Ascending `event_state`
    - Descending `pct_white_zip`
    - Descending `med_inc`
    
    Then preview the first 10 rows using `head()`. Do this using `tidyverse`'s `arrange()` function.

__tidyverse using `arrange()`__:
```{r}
df_event <- arrange(df_event, univ_id, event_date, event_state, desc(pct_white_zip), desc(med_inc))
head(df_event, n = 10)
```

# Question 4: Selecting variables


1. Create a new dataframe by selecting the columns `univ_id`, `event_date`, `event_type`, `zip`, and `med_inc` from `df_event`. Use the `names()` function to show what columns (variables) are in the newly created dataframe.

    Do this in 2 ways: using `tidyverse`'s `select()` and `base R`'s subsetting operators.

__tidyverse using `select()`__:
```{r}
df_event_tidy <- select(df_event, univ_id, event_date, event_type, zip, med_inc)
names(df_event_tidy)
```

__base R using subsetting operators__:
```{r}
df_event_R <- df_event[c(2,5,6,7,11)]
names(df_event_R)
```


# Question 5: Additional practice with `df_school_all` dataframe


1. In the code chunk below, complete the following:

    - Use the `load()` and `url()` functions to download the `df_school_all` dataframe from the url: `https://github.com/anyone-can-cook/rclass1/raw/master/data/recruiting/recruit_school_allvars.RData`
      - Each row in `df_school_all` represents a high school (includes both public and private)
      - There are columns (e.g., `visit_by_100751`) indicating the number of times a university visited that high school
      - The variable `total_visits` identifies the number of visits the high school received from all (16) public research universities in this data collection sample
    - Use `names()` to identify the column names in the dataframe
    - Use `table()` to show the categorical values of the `school_type` column

```{r}
load(url("https://github.com/anyone-can-cook/rclass1/raw/master/data/recruiting/recruit_school_allvars.RData"))
str(df_school_all)
names(df_school_all)
table(df_school_all$school_type)
```


2. Use the tidyverse functions `arrange()` and `select()` to do the following:

    - Sort `df_school_all` descending by `total_visits`
    - Select the following variables: `name`, `state_code`, `city`, `school_type`,`total_visits`, `med_inc`, `pct_white`, `pct_black`, `pct_hispanic`, `pct_asian`, `pct_amerindian`
      - Note: You can do this in one step by wrapping the `select()` function around `arrange()`, or you can do this in two steps by creating an intermediate dataframe.
      
    Print the first 10 rows of the final dataframe using `head()`, which represents the top 10 most visited schools by the 16 universities.

```{r}
df_school_all <- select(arrange(df_school_all, desc(total_visits)), name, state_code, city, school_type, total_visits, med_inc, pct_white, pct_black, pct_hispanic, pct_asian, pct_amerindian)
head(df_school_all, n = 10)
```


3. Building upon the previous question, print the following (select same variables as above):  

    (A) Top 10 most visited public high schools in California
    (B) Top 10 most visited private high schoools in California
    
```{r}
#the arrangement desc(total_visits)) was assigned to the dataframe df_school_all when completing question 5.2. 

filter(df_school_all, school_type == "public", state_code == "CA") %>% head(df_school_all, n = 10)
filter(df_school_all, school_type == "private", state_code == "CA") %>% head(df_school_all, n = 10)
```

# Create a GitHub issue   

- Go to the [class repository](https://github.com/anyone-can-cook/rclass1_student_issues_f23/issues) and create a new issue.

- Refer to [rclass1 student issues readme](https://github.com/anyone-can-cook/rclass1_student_issues_f23/blob/main/README.md) for instructions on how to post questions or reflections.

- You are also required to respond to at least one issue posted by another student.

- Paste the url to your issue here: https://github.com/anyone-can-cook/rclass1_student_issues_f23/issues/399 
- Paste the url to the issue you responded to here: https://github.com/anyone-can-cook/rclass1_student_issues_f23/issues/398 

# Knit to pdf and submit problem set  

getwd()
setwd("C:/Users/there/OneDrive/Documents)

**Knit to pdf** by clicking the "Knit" button near the top of your RStudio window (icon with blue yarn ball) or drop down and select "Knit to PDF"

- Go to the [class website](https://anyone-can-cook.github.io/rclass1/) and under the "Readings & Assignments" >> "Week 3" tab, click on the "Problem set 3 submission link"  
- Submit both .Rmd and pdf files  
- Use this naming convention "lastname_firstname_ps#" for your .Rmd and pdf files (e.g. jaquette_ozan_ps3.Rmd & jaquette_ozan_ps3.pdf)  
